Staged Compilation with Two-Level Type Theory

### 2022-07-29 paper/paper.tex
edits

\paragraph{Closed modalities}\label{sec:closed-modalities}
Another option is to add a \emph{closed} or \emph{crisp} modality
\cite{licata2018internal} for tracking object-level values which will become
closed terms after staging. Since closed terms are not affected by substitution,
we are free to analyze them. Such analysis can be viewed as an internalization
of canonicity for closed object-level terms. For example, a closed $\Bool_0$
term will be either $\true_0$ or $\false_0$ after staging, so it makes sense to
be able to eliminate from it to $\Bool_1$. This can be also viewed as a safe
implementation of a \emph{run} function in the style of MetaML \cite{metaml}.
In our case \emph{scope extrusion} would be prevented by the closed modality.

\emph{Function pointers} in the style of the C programming language could be also
an interesting use case, since these must become closed after staging (as they
cannot capture closure environments).

@@

The multi-stage calculus in \cite{multi-stage-calculus} supports staging
operations and type dependencies. However, there are major differences to the
current work. First, ibid.\ does not have large elimination, hence it does not
support computing types by staging. Second, it does not support staged
compilation in the sense of Definition \ref{def:staging}; rather, it is a system
for runtime code manipulation and evaluation (which we may call staged
\emph{computation}). In other words, programs may manipulate code fragments
during execution, in more liberal ways than in 2LTT, but there is no judgment
which ensures that a purely ``object-level'' program can be extracted from a
multi-stage program. Similar considerations apply to systems with the S4
modality (see e.g.\ \cite{DBLP:journals/jacm/DaviesP01}) where there is no
judgment that guarantees elimination of modal operations.

@@

In MetaML and MetaOCaml, there is a \emph{run} function for evaluating closed
object-level terms: as we mentioned in Section \ref{sec:closed-modalities}, we
believe that the proper way to implement it in 2LTT would be to use a closed
modality. This use case is closely related to \cite{licata2018internal}. In
ibid.\ a semantics is sketched for this modality in a presheaf topos, and we
could directly adapt it to our presheaf model of 2LTT. That said, working
out the practical details of this modality remains future work.

>>>>

\paragraph{Closed modalities}\label{sec:closed-modalities}
Another option is to add a \emph{closed} or \emph{crisp} modality for modeling
closed object-level terms \cite{licata2018internal}. Since closed terms are not
affected by substitution, we are able to analyze them, and we can also recover
open terms by explicitly abstracting over contexts
\cite{DBLP:conf/lics/Hofmann99,DBLP:journals/corr/abs-2206-02831}. \emph{Function
pointers} in the style of the C programming language could be also an
interesting use case, since these must become closed after staging (as they
cannot capture closure environments).

%% for tracking object-level values which will become
%% closed terms after staging. Since closed terms are not affected by substitution,
%% we are free to analyze them. Such analysis can be viewed as an internalization
%% of canonicity for closed object-level terms. For example, a closed $\Bool_0$
%% term will be either $\true_0$ or $\false_0$ after staging, so it makes sense to
%% be able to eliminate from it to $\Bool_1$. This can be also viewed as a safe
%% implementation of a \emph{run} function in the style of MetaML \cite{metaml}.
%% In our case \emph{scope extrusion} would be prevented by the closed modality.

@@

In \cite{DBLP:conf/lics/Hofmann99}, it is described how to model higher-order
abstract syntax of simply typed object languages in presheaves over various
syntactic categories. In the case where syntactic morphisms are substitutions,
the internal language described by Hofmann can be viewed as an alternative
syntax for a 2LTT, where we only have a single family structure and the object
level is modeled with a Tarski-style universe.

The multi-stage calculus in \cite{multi-stage-calculus} supports staging
operations and type dependencies. However, there are major differences to the
current work. First, ibid.\ does not have large elimination, hence it does not
support computing types by staging. Second, it does not support staged
compilation in the sense of Definition \ref{def:staging}; there is no judgment
which ensures that staging operations can be eliminated and a purely
``object-level'' program can be extracted from a multi-stage program. Rather, the system
supports \emph{runtime} code generation. There are several modal multi-stage
systems which support runtime code generation but not staged compilation, such
as those based on S4 modalities (e.g.\ \cite{DBLP:journals/jacm/DaviesP01}) or
contextual modalities (\cite{DBLP:journals/pacmpl/JangGMP22}).

@@

In MetaOCaml, there is a \emph{run} function for evaluating closed object-level
terms, however, this can fail during staging since closedness is not statically
ensured. In our setting, this could be reproduced in a safe way using a closed
modality; this is part of future work.

%% as we mentioned in Section \ref{sec:closed-modalities}, we
%% believe that the proper way to implement it in 2LTT would be to use a closed
%% modality. This use case is closely related to \cite{licata2018internal}. In
%% ibid.\ a semantics is sketched for this modality in a presheaf topos, and we
%% could directly adapt it to our presheaf model of 2LTT. That said, working
%% out the practical details of this modality remains future work.
